Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2017-08-10 threshold/draft.Rnw
Wordsmiting and reorganization.

%\Abstract{
\begin{document}
\maketitle
\begin{abstract}
Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.
\end{abstract}
%\Keywords{thresholded correlation, SVD, \proglang{R}}
%\Plainkeywords{thresholded correlation, SVD, R}
%\Address{
%James Baglama, Chair of Mathematics\\
%Department of Mathematics\\
%University of Rhode Island\\
%Kingston, Rhode Island 02881\\
%E-mail: \email{jbaglama@math.uri.edu}\\
%URL: \url{http://www.math.uri.edu/~jbaglama}\\
%\\
%Michael Kane\\
%Yale University\\
%300 Geroge Street\\
%New Haven, Connecticut 06511\\
%E-mail: \email{michael.kane@yale.edu}\\
%\\
%Bryan Lewis\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{blewis@illposed.net}\\
%URL: \url{http://illposed.net}\\
%\\
%Alex Poliakov\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{apoliakov@paradigm4.com}\\
%}

@@

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes \cite{genomics}.  Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications \cite{ml1, ml2} and are important to time series and
image processing applications \cite{timeseries, svd-similarity}.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning methods} that cheaply
prune away all but the most highly-correlated pairs of vectors.  For example
the TAPER and related algorithms of Xiong et al. develop pruning rules based on
the frequency of mutually co-occurring observations (rows) among vector pairs;
see \cite{prune1, prune2} and the references therein.  Related methods
approximate a set containing the most highly-correlated pairs using hashing
algorithms \cite{prune3}. A number of other methods arrive at approximate sets
of the most highly-correlated pairs using dimensionality-reduction techniques
like the discrete Fourier transform \cite{timeseries}. The method of Wu, et al.
\cite{svd-similarity}, conceptually very similar to our approach, uses the
singular value decomposition (SVD) to find all pairs of close vectors with
respect to a distance metric (instead of correlation). Hero and
Rajaratnam\cite{hero2011large} apply projections to the thresholded correlation
problem with the additional objective of determining a reasonable threshold
value; their method follows a probabilistic approach.  Our method finds all of
the most highly-correlated vector pairs with respect to a given threshold by
pruning along an intuitive path of decreasing variance defined by the data and
computed by a truncated SVD.

@@

\section{Efficient pruning using truncated SVD}

@@

@@

We have so far assumed that the mean of each column vector of the $m\times n$
matrix $A$ is zero and the columns are scaled to have unit norm, but we really
want a method that works with general matrices $A$.  We also want a way to
efficiently compute a relatively low-rank truncated SVD of potentially large
matrices required by step 1 of Algorithm \ref{prune}.  And the truncated SVD
method should be able to cheaply restart to compute additional singular vectors
as required in step 5 of Algorithm \ref{prune}. Fortunately all desires are met
by one algorithm.

The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for \proglang{R} \cite{irlbar}, \proglang{Matlab} \cite{irlbam},
and \proglang{Python} \cite{irlbap}.

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without any constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

@@

\section{Practical extensions}\label{extensions}

>>>>

%\Abstract{
\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle
\begin{abstract}
Efficiently computing a subset of a correlation matrix consisting of values above 
a specified threshold is important to many practical applications of large-scale 
statistics and machine learning. Real-world problems in genomics, machine 
learning, finance, and others can produce correlation matrices too large to 
explicitly form and to tractably compute. However, in most applications, only 
values corresponding to highly-correlated (or anti-correlated) vectors are of 
interest, and those corresponding values typically make up a small fraction of 
the overall correlation matrix. For this situation, we present a method based on 
the singular value decomposition (SVD) and its relationship to the data 
covariance structure for efficiently computing thresholded subsets of very large correlation matrices.
\end{abstract}
%\Keywords{thresholded correlation, SVD, \proglang{R}}
%\Plainkeywords{thresholded correlation, SVD, R}
%\Address{
%James Baglama, Chair of Mathematics\\
%Department of Mathematics\\
%University of Rhode Island\\
%Kingston, Rhode Island 02881\\
%E-mail: \email{jbaglama@math.uri.edu}\\
%URL: \url{http://www.math.uri.edu/~jbaglama}\\
%\\
%Michael Kane\\
%Yale University\\
%300 Geroge Street\\
%New Haven, Connecticut 06511\\
%E-mail: \email{michael.kane@yale.edu}\\
%\\
%Bryan Lewis\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{blewis@illposed.net}\\
%URL: \url{http://illposed.net}\\
%\\
%Alex Poliakov\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{apoliakov@paradigm4.com}\\
%}

@@

Finding highly-correlated pairs among a large set of vectors is an important part 
of many statistics and machine learning applications. For instance, subsets of 
highly-correlated vectors of gene expression values may be used in the discovery 
of networks of genes relevant to particular biological processes \cite{genomics}. 
Identification of highly-correlated pairs may be used in feature selection 
algorithms for model fitting \cite{ml1, ml2} and are important to time series and 
image processing applications \cite{timeseries, svd-similarity}.

Because the number of correlation coefficients grows quadratically with the 
number of vectors, simply computing all pairs of correlation coefficients and 
then filtering out coefficients below a given threshold may not be 
computationally feasible for large-scale data applications. As a result, 
considerable attention has been devoted to \emph{pruning methods} that cheaply 
prune away all but the most highly-correlated pairs of vectors.  For example the 
TAPER and related algorithms of Xiong et al. develop pruning rules based on the 
frequency of mutually co-occurring observations (rows) among vector pairs; see 
\cite{prune1, prune2} and the references therein.  Related methods approximate a 
set containing the most highly-correlated pairs using hashing algorithms 
\cite{prune3}. Other methods arrive at approximate sets of the most 
highly-correlated pairs using dimensionality-reduction techniques like the 
discrete Fourier transform \cite{timeseries}.

The method of Wu, et al. \cite{svd-similarity}, conceptually very similar to our 
approach, uses the singular value decomposition (SVD) to find all pairs of close 
vectors with respect to a distance metric (instead of correlation). Similarly, 
Hero and Rajaratnam\cite{hero2011large} apply projections to the thresholded 
correlation problem with the additional objective of determining a reasonable 
threshold value; their method follows a probabilistic approach.  Our method finds 
all of the most highly-correlated vector pairs with respect to a given threshold 
by pruning along an intuitive path of decreasing variance defined by the data and
computed by a truncated SVD.

@@

\section{Efficient pruning using truncated SVD} \label{sect:algo}

@@

\section{Extensions}\label{extensions}

@@

For Algorithm \ref{prune} to be used in practice, three challenges need to be 
addressed. First, the algorithm requires an efficient method for computing a 
relatively low-rank truncated SVD of potentially large matrices required by step 
1. Second, the truncated SVD method should be able to cheaply restart to compute 
additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
Third, while the algorithm assumes a standardized matrix, it should work for 
general matrices. Fortunately, each of these challenges is met with a single 
algorithm.

The first two challenges are addressed directly by the augmented implicitly 
restarted Lanczos bidiagonalization algorithm\break(IRLBA) of Baglama and Reichel 
\cite{irlba}. It efficiently computes truncated singular value decompositions of 
either large dense or sparse matrices and the result from the $k$ rank approximati
on can be efficiently used to create an approximation with higher rank. Reference 
implementations are available for \proglang{R} \cite{irlbar}, \proglang{Matlab} 
\cite{irlbam}, and \proglang{Python} \cite{irlbap}.

The third challenge is addressed by relaxing the column mean and scale 
assumptions without introducing significant additional computational or storage
overhead as follows. Assume $A$ is an $m\times n$ real-valued matrix without any 
constant-valued columns. Let $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent 
the vector of column means of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
%The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
%products. The main idea behind efficient application of IRLB to correlation
%problems replaces 
Matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ may be used in
the iterations of the algrithm, implicitly working with a scaled and 
centered matrix without explicit standardization. This comes at the cost 
of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and scaling the input matrix.

@@

### 2017-08-10 threshold/draft.tex
Wordsmiting and reorganization.

%\Abstract{
\begin{document}
\maketitle
\begin{abstract}
Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.
\end{abstract}
%\Keywords{thresholded correlation, SVD, \proglang{R}}
%\Plainkeywords{thresholded correlation, SVD, R}
%\Address{
%James Baglama, Chair of Mathematics\\
%Department of Mathematics\\
%University of Rhode Island\\
%Kingston, Rhode Island 02881\\
%E-mail: \email{jbaglama@math.uri.edu}\\
%URL: \url{http://www.math.uri.edu/~jbaglama}\\
%\\
%Michael Kane\\
%Yale University\\
%300 Geroge Street\\
%New Haven, Connecticut 06511\\
%E-mail: \email{michael.kane@yale.edu}\\
%\\
%Bryan Lewis\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{blewis@illposed.net}\\
%URL: \url{http://illposed.net}\\
%\\
%Alex Poliakov\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{apoliakov@paradigm4.com}\\
%}

@@

@@

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:12:12 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.2 by xtable 1.8-2 package
% Sat Sep  3 20:12:12 2016
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

\section{Practical extensions}\label{extensions}

Arbitrary vectors

>>>>

%\Abstract{
\begin{document}
\input{draft-concordance}
\maketitle
\begin{abstract}
Efficiently computing a subset of a correlation matrix consisting of values above 
a specified threshold is important to many practical applications of large-scale 
statistics and machine learning. Real-world problems in genomics, machine 
learning, finance, and others can produce correlation matrices too large to 
explicitly form and to tractably compute. However, in most applications, only 
values corresponding to highly-correlated (or anti-correlated) vectors are of 
interest, and those corresponding values typically make up a small fraction of 
the overall correlation matrix. For this situation, we present a method based on 
the singular value decomposition (SVD) and its relationship to the data 
covariance structure for efficiently computing thresholded subsets of very large correlation matrices.
\end{abstract}
%\Keywords{thresholded correlation, SVD, \proglang{R}}
%\Plainkeywords{thresholded correlation, SVD, R}
%\Address{
%James Baglama, Chair of Mathematics\\
%Department of Mathematics\\
%University of Rhode Island\\
%Kingston, Rhode Island 02881\\
%E-mail: \email{jbaglama@math.uri.edu}\\
%URL: \url{http://www.math.uri.edu/~jbaglama}\\
%\\
%Michael Kane\\
%Yale University\\
%300 Geroge Street\\
%New Haven, Connecticut 06511\\
%E-mail: \email{michael.kane@yale.edu}\\
%\\
%Bryan Lewis\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{blewis@illposed.net}\\
%URL: \url{http://illposed.net}\\
%\\
%Alex Poliakov\\
%Paradigm4, Inc.\\
%281 Winter Street, Suite 360\\
%Waltham, MA 02451\\
%E-mail: \email{apoliakov@paradigm4.com}\\
%}

@@

\section{Extensions}\label{extensions}

WRITE ME

@@

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Thu Aug 10 12:49:00 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
\endgroup
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Thu Aug 10 12:49:00 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
\endgroup
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.
